---
title: "Targeted peak integration of LC-MS data using T.A.R.D.I.S."
author: "Pablo Vangeenderhuysen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Targeted peak integration of LC-MS data using T.A.R.D.I.S.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(TARDIS)
```


# Introduction

`T.A.R.D.I.S.` is an R package developed by Pablo Vangeenderhuysen at LIMET, Ghent University. 
`T.A.R.D.I.S.` offers an easy and straightforward way to automatically calculate area under the peak, max intensity and various quality metrics for targeted chemical compounds in LC-MS data. It makes use of established retention time correction algorithms from the `xcms` package and loads MS data as `Spectra`objects so it's easily integrated with other tools of the *Rformassspectrometry* initiative.

See README for installation instructions.

This tutorial explains the basic functionalities of `T.A.R.D.I.S.`and how to use the Shiny GUI.

# Using the graphical user interface of T.A.R.D.I.S.

To launch the GUI run:

```{r eval=FALSE, include=TRUE}
runTardis()
```



