---
title: "Testing different targeted analysis methods"
author: "Pvgeende"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document was constructed to test different methods to extract and integrate peaks in LC-MS data. Showing the need for a better integrated and better performing tool. As a quick test and proof of concept, two runs of an analytical standard mixture are used. If everything goes well, these concepts, scripts and functions can be directly applied to the final sample datasets.

```{r include=FALSE}
library(xcms)
library(Spectra)
library(MsExperiment)
library(RColorBrewer)
library(peakPantheR)
library(SummarizedExperiment)
library(MetaboAnnotation)
library(pracma)
library(signal)
library(tidyverse)
library(stringr)
library(readxl)
library(Polychrome)
library(openxlsx)

```


## Prepare data & files 

### Loading spectra data

I use two runs of a mixture of analytical standards. Mix1: 25 analytes in positive ionisation.
Using Spectra to load data.

```{r}

files <- list.files("K:/shares/di04_limet_bioinformatics/PhD Pablo/Tapex_test_STD_Environage/pos",full.names = T)
files <- files[c(1,11)]
data <- readMsExperiment(spectraFiles = files)

```


### Get compound info

I will be using a 5ppm search window on m/z and 2min search window for RT (better for XCMS).

```{r}

masslist <- read_excel("K:/shares/di04_limet_bioinformatics/PhD Pablo/Tapex_test_STD_Environage/targlijst.xlsx")
masslist <- masslist[1:34,] #ONLY COMP FOR MIX 1
polarity <- "positive" #you can change the polarity here, needs to be "positive" or "negative"
mode <- "metabolomics" #you can change the mode here, needs to be "lipidomics" or "metabolomics"
#With these two parameters, you can set the allowed errors for mass & retention time!
ppm <- 10 #ppm error for EIC extraction #POLAR = 5ppm / LIPIDOMICS =  10 ppm
deltaTR = 120 

masslist_negative <- masslist[grep("]-",masslist$`Ion adduct`,fixed = T),]
masslist_positive <- masslist[grep("]+",masslist$`Ion adduct`,fixed = T),]

#Remove unnecessary columns and rename

masslist_positive <- masslist_positive[,c("ID","Name","m/z-value","RT (min)")]
masslist_negative <- masslist_negative[,c("ID","Name","m/z-value","RT (min)")]
colnames(masslist_positive) <- c("ID","NAME","m/z","tr")
colnames(masslist_negative) <- c("ID","NAME","m/z","tr")

#Set RT in seconds & make numeric

masslist_positive$tr <- as.numeric(masslist_positive$tr) *60
masslist_positive$`m/z` <- as.numeric(masslist_positive$'m/z')

masslist_negative$tr <- as.numeric(masslist_negative$tr) *60
masslist_negative$`m/z` <- as.numeric(masslist_negative$'m/z')

#set ID as character

masslist_positive$ID <- as.character(masslist_positive$ID)
masslist_negative$ID <- as.character(masslist_negative$ID)

if(polarity == "positive"){
  dbData <- masslist_positive
}
if (polarity == "negative"){
  dbData <- masslist_negative
}

mzmed <- dbData$`m/z`
rtmed <- dbData$tr



mzdeltas <- sapply(mzmed, function(mzmed) mzmed*ppm/10^6)

#calculate mzrange based on delta mz
mzRanges <- cbind(as.numeric(mzmed) - mzdeltas, as.numeric(mzmed) + mzdeltas )

spectra = data@spectra

#if an upper m/z boundary is lower than the minimum m/z range, set it to the minimum m/z
indexTemp <- which(mzRanges[,2] < min(spectra@backend@spectraData@listData$basePeakMZ))
mzRanges[indexTemp,] <- min(spectra@backend@spectraData@listData$basePeakMZ)

#if a lower m/z boundary is higher than the max m/z, set it to the max m/z
indexTemp <- which(mzRanges[,1] > max(spectra@backend@spectraData@listData$basePeakMZ))
mzRanges[indexTemp,] <- max(spectra@backend@spectraData@listData$basePeakMZ)

#if an upper limit is higher than the max mz & the lower limit is smaller than the max m/z, set the upper limit to the max m/z
mzRanges[which(mzRanges[,2] > max(spectra@backend@spectraData@listData$basePeakMZ) & mzRanges[,1] < max(spectra@backend@spectraData@listData$basePeakMZ)),2] <- max(spectra@backend@spectraData@listData$basePeakMZ)

#if a upper limit is larger than the minimum & the lower limit is lower than the minimum, set the lower limit to the min m/z
mzRanges[which(mzRanges[,2] > min(spectra@backend@spectraData@listData$basePeakMZ) & mzRanges[,1] < min(spectra@backend@spectraData@listData$basePeakMZ)),1] <- min(spectra@backend@spectraData@listData$basePeakMZ)



rtRanges <- cbind(as.numeric(rtmed) - deltaTR/2, as.numeric(rtmed) + deltaTR/2)

#for al upper RT limits -2 lower than minimum RT, change to lower limit of min rt & upper limit to min + 10
indexTemp <- which(rtRanges[,2] - 2 < min(spectra@backend@spectraData@listData$rtime))
rtRanges[indexTemp,] <- cbind(rep(min(spectra@backend@spectraData@listData$rtime),length(indexTemp)), rep(min(spectra@backend@spectraData@listData$rtime)+10,length(indexTemp)))


#for al lower RT limits +2 higher than maximum RT, change lower limit to max -10 and upper limit to max
indexTemp <- which(rtRanges[,1] + 2 > max(spectra@backend@spectraData@listData$rtime))
rtRanges[indexTemp,] <- cbind(rep(max(spectra@backend@spectraData@listData$rtime)-10,length(indexTemp)), rep(max(spectra@backend@spectraData@listData$rtime),length(indexTemp)))


#for upper limits higher than max and lower limit lower than max, change upper limit to max
rtRanges[which(rtRanges[,2] > max(spectra@backend@spectraData@listData$rtime) & rtRanges[,1] < max(spectra@backend@spectraData@listData$rtime)),2] <- max(spectra@backend@spectraData@listData$rtime)
#for upper limits higher than min and lower limit lower than min, change lower limit to min
rtRanges[which(rtRanges[,2] > min(spectra@backend@spectraData@listData$rtime) & rtRanges[,1] < min(spectra@backend@spectraData@listData$rtime)),1] <- min(spectra@backend@spectraData@listData$rtime)

```

### Inspect data

Take a look at the chromatograms. Both chromatograms look good, although TIC intensity seems to differ quite a lot between both runs of the analytical standard mixture.

```{r}
group_colors <- paste0(brewer.pal(3, "Set1")[1:2], "60")
names(group_colors) <- c("1", "2")
bpis <- chromatogram(data, aggregationFun = "sum")
plot(bpis, col = group_colors, main = "TIC Chromatogram")   
legend(800, 1.5E9, legend=c("Run 1", "Run 2"),
       col= group_colors, lty=1:2, cex=0.8)

```

## XCMS centWave: untargeted approach

Here I first try to detect all the targeted compounds after running the "default" xcms workflow for untargeted processing.
For the parameters, I used approx. the same settings as used in the "picky with peakpicking" paper by William Kumler and in the xcms vignette. I lowered min peakwidth from 20 to 10. 

First peak detection:

```{r}
cwp <- CentWaveParam(ppm = 5, peakwidth = c(10,80), integrate = 2, snthresh = 0, prefilter = c(3,10e6), extendLengthMSW = T, verboseColumns = T )

xdata <- findChromPeaks(data, param = cwp)
```

Merging neighbouhring peaks:

```{r}
mpp <- MergeNeighboringPeaksParam(expandRt = 4)
xdata_pp <- refineChromPeaks(xdata, mpp)
xdata <- xdata_pp

```

RT alignment, also reduced binSize to 0.1 as Kulner.

```{r}
xdata <- adjustRtime(xdata, param = ObiwarpParam(binSize = 0.1))
```
Peak grouping:

```{r}
pdp <- PeakDensityParam(sampleGroups = c("STD","STD"),
                        minFraction = 0.5, bw = 12, minSamples = 1, binSize = 0.001)
xdata <- groupChromPeaks(xdata, param = pdp)
```

Gap filling:

```{r}
xdata <- fillChromPeaks(xdata, param = ChromPeakAreaParam())

```

Finalize results:

```{r}
res <- quantify(xdata, value = "into", method = "sum")
```


```{r}
rowData(res)
```

Area's of the features in both samples can also be seen:

```{r}
featureValues(xdata)
```

We end up with 293 features, now we will use the MetaboAnnotation package to see if we can retrieve the targeted analytes from our mixture.

```{r}
ms1feat <- data.frame(mz = rowData(res)$mz,rt = rowData(res)$rt)
target_db <- data.frame(name = masslist_positive$NAME, rt = masslist_positive$tr, mz = masslist_positive$`m/z`)

prm <- MzRtParam(ppm = 10, toleranceRt = 20)
mtch <- matchValues(ms1feat, target_db,  param = prm,
                    mzColname = c("mz", "mz"), rtColname = c("rt", "rt"))
mtch
```

19 of the 293 MS1 features are matched with 20 of the 25 target objects. So one MS1 feature got matched with two target analytes.
In this case, feature number 138 got matched to both (2S,3S)-2-amino-3-methylpentanoic acid (m/z = 132.10191 and rt = 128.4) and (2S)-2-amino-4-methylpentanoic acid (m/z = 132.10191 and 142.2).

In detail:

```{r}
rowData(res)[whichQuery(mtch),]
```

All of them are present in both of the runs.




Take a look and save their chromatograms for manual inspection. --> they all look good!

```{r}
chromatograms <- featureChromatograms(xdata)

found_feat <- chromatograms[whichQuery(mtch)]
```




```{r}
save_location <- "K:/shares/di04_limet_bioinformatics/PhD Pablo/Publicaties/WIP/TARDIS/figures/proof_of_concept/centwave_untargeted_plots_mix_1/"

df = matches(mtch)
names <- c()

for(i in 1:19){
  name <- masslist_positive$NAME[df$target_idx[df$query_idx == df$query_idx[i]]]
  names <- c(names,name)
}

names <- unique(names)

for(i in 1:length(whichQuery(mtch))){
  png(filename = paste0(save_location,names[i],".png"))
  plot(found_feat[i],main = names[i])
  dev.off()
}
  

```


## XCMS centWave: targeted approach



```{r}
results = NULL
cwp <- CentWaveParam(ppm = 5, peakwidth = c(10,80), integrate = 2, snthresh = 0, prefilter = c(3,10e6), extendLengthMSW = T, verboseColumns = T )


for(compound in 1:25){
  mzr <- mzRanges[compound,]
  rtr <- rtRanges[compound,]
  
  xdata <- data |>
    filterRt(rt = rtr) |>
    filterMz(mz = mzr) |>
    findChromPeaks(param = cwp)
  
  if(isEmpty(chromPeaks(xdata)) == TRUE){
    result = c(0,0)
    
  } else{
    
    
    xdata <- adjustRtime(xdata, param = ObiwarpParam(binSize = 0.1))
    
    pdp <- PeakDensityParam(sampleGroups = sampleData(xdata)$spectraOrigin,
                            minFraction = 0.5, bw = 12, minSamples = 1, binSize = 0.001)
    xdata <- groupChromPeaks(xdata, param = pdp)
    
    result <- featureValues(xdata, value = "into")
  }
  
   
    
    results <- rbind(results,result)
}
```

Very few features are detected using centWave with the same parameters on a smaller m/z & RT window. 

```{r}
kableExtra::kable(results)
```

Trying again with a different set of parameters (changing prefilter to the default c(3,100)):

```{r}
results = NULL
cwp <- CentWaveParam(ppm = 5, peakwidth = c(10,80), integrate = 2, snthresh = 0, extendLengthMSW = T, verboseColumns = T)


for(compound in 1:25){
  mzr <- mzRanges[compound,]
  rtr <- rtRanges[compound,]
  
  xdata <- data |>
    filterRt(rt = rtr) |>
    filterMz(mz = mzr) |>
    findChromPeaks(param = cwp)
  
  if(isEmpty(chromPeaks(xdata)) == TRUE){
    result = c(0,0)
    
  } else if(length(unique(chromPeaks(xdata)[,"sample"])) >= 2){
    xdata <- adjustRtime(xdata, param = ObiwarpParam(binSize = 0.1))
     pdp <- PeakDensityParam(sampleGroups = sampleData(xdata)$spectraOrigin,
                            minFraction = 0.5, bw = 12, minSamples = 1, binSize = 0.001)
    xdata <- groupChromPeaks(xdata, param = pdp)
    
    result <- featureValues(xdata, value = "into")
  }
  
  else{
    
    
    
    pdp <- PeakDensityParam(sampleGroups = sampleData(xdata)$spectraOrigin,
                            minFraction = 0.5, bw = 12, minSamples = 1, binSize = 0.001)
    xdata <- groupChromPeaks(xdata, param = pdp)
    
    result <- featureValues(xdata, value = "into")
  }
  
   
    
    results <- rbind(results,result)
}
```

Not much of an improvement, also quite some times where multiple features are detected in the filtered chromatograms (FT2 rows)

```{r}
kableExtra::kable(results)
```


## peakPantheR

Workflow as described in the parallel annotation vignette: https://bioconductor.org/packages/release/bioc/vignettes/peakPantheR/inst/doc/parallel-annotation.html

First redefine the rt borders to be smaller:

```{r}
library(readxl)
masslist <- read_excel("K:/shares/di04_limet_bioinformatics/PhD Pablo/Tapex_test_STD_Environage/targlijst.xlsx")
masslist <- masslist[1:34,] #ONLY COMP FOR MIX 1
polarity <- "positive" #you can change the polarity here, needs to be "positive" or "negative"
mode <- "metabolomics" #you can change the mode here, needs to be "lipidomics" or "metabolomics"
#With these two parameters, you can set the allowed errors for mass & retention time!
ppm <- 10 #ppm error for EIC extraction #POLAR = 5ppm / LIPIDOMICS =  10 ppm
deltaTR = 18 

masslist_negative <- masslist[grep("]-",masslist$`Ion adduct`,fixed = T),]
masslist_positive <- masslist[grep("]+",masslist$`Ion adduct`,fixed = T),]

#Remove unnecessary columns and rename

masslist_positive <- masslist_positive[,c("ID","Name","m/z-value","RT (min)")]
masslist_negative <- masslist_negative[,c("ID","Name","m/z-value","RT (min)")]
colnames(masslist_positive) <- c("ID","NAME","m/z","tr")
colnames(masslist_negative) <- c("ID","NAME","m/z","tr")

#Set RT in seconds & make numeric

masslist_positive$tr <- as.numeric(masslist_positive$tr) *60
masslist_positive$`m/z` <- as.numeric(masslist_positive$'m/z')

masslist_negative$tr <- as.numeric(masslist_negative$tr) *60
masslist_negative$`m/z` <- as.numeric(masslist_negative$'m/z')

#set ID as character

masslist_positive$ID <- as.character(masslist_positive$ID)
masslist_negative$ID <- as.character(masslist_negative$ID)

if(polarity == "positive"){
  dbData <- masslist_positive
}
if (polarity == "negative"){
  dbData <- masslist_negative
}

mzmed <- dbData$`m/z`
rtmed <- dbData$tr



mzdeltas <- sapply(mzmed, function(mzmed) mzmed*ppm/10^6)

#calculate mzrange based on delta mz
mzRanges <- cbind(as.numeric(mzmed) - mzdeltas, as.numeric(mzmed) + mzdeltas )

spectra = data@spectra

#if an upper m/z boundary is lower than the minimum m/z range, set it to the minimum m/z
indexTemp <- which(mzRanges[,2] < min(spectra@backend@spectraData@listData$basePeakMZ))
mzRanges[indexTemp,] <- min(spectra@backend@spectraData@listData$basePeakMZ)

#if a lower m/z boundary is higher than the max m/z, set it to the max m/z
indexTemp <- which(mzRanges[,1] > max(spectra@backend@spectraData@listData$basePeakMZ))
mzRanges[indexTemp,] <- max(spectra@backend@spectraData@listData$basePeakMZ)

#if an upper limit is higher than the max mz & the lower limit is smaller than the max m/z, set the upper limit to the max m/z
mzRanges[which(mzRanges[,2] > max(spectra@backend@spectraData@listData$basePeakMZ) & mzRanges[,1] < max(spectra@backend@spectraData@listData$basePeakMZ)),2] <- max(spectra@backend@spectraData@listData$basePeakMZ)

#if a upper limit is larger than the minimum & the lower limit is lower than the minimum, set the lower limit to the min m/z
mzRanges[which(mzRanges[,2] > min(spectra@backend@spectraData@listData$basePeakMZ) & mzRanges[,1] < min(spectra@backend@spectraData@listData$basePeakMZ)),1] <- min(spectra@backend@spectraData@listData$basePeakMZ)



rtRanges <- cbind(as.numeric(rtmed) - deltaTR/2, as.numeric(rtmed) + deltaTR/2)

#for al upper RT limits -2 lower than minimum RT, change to lower limit of min rt & upper limit to min + 10
indexTemp <- which(rtRanges[,2] - 2 < min(spectra@backend@spectraData@listData$rtime))
rtRanges[indexTemp,] <- cbind(rep(min(spectra@backend@spectraData@listData$rtime),length(indexTemp)), rep(min(spectra@backend@spectraData@listData$rtime)+10,length(indexTemp)))


#for al lower RT limits +2 higher than maximum RT, change lower limit to max -10 and upper limit to max
indexTemp <- which(rtRanges[,1] + 2 > max(spectra@backend@spectraData@listData$rtime))
rtRanges[indexTemp,] <- cbind(rep(max(spectra@backend@spectraData@listData$rtime)-10,length(indexTemp)), rep(max(spectra@backend@spectraData@listData$rtime),length(indexTemp)))


#for upper limits higher than max and lower limit lower than max, change upper limit to max
rtRanges[which(rtRanges[,2] > max(spectra@backend@spectraData@listData$rtime) & rtRanges[,1] < max(spectra@backend@spectraData@listData$rtime)),2] <- max(spectra@backend@spectraData@listData$rtime)
#for upper limits higher than min and lower limit lower than min, change lower limit to min
rtRanges[which(rtRanges[,2] > min(spectra@backend@spectraData@listData$rtime) & rtRanges[,1] < min(spectra@backend@spectraData@listData$rtime)),1] <- min(spectra@backend@spectraData@listData$rtime)

```




```{r}
input_spectraPaths <- files
input_targettable <- masslist_positive
colnames(input_targettable) <- c("cpdID","cpdName","mz","rt")
input_targettable$rtMin <- rtRanges[,1]
input_targettable$rtMax <- rtRanges[,2]
input_targettable$mzMin <- mzRanges[,1]
input_targettable$mzMax <- mzRanges[,2]

```

```{r}
init_annotation <- peakPantheRAnnotation(spectraPaths = input_spectraPaths,
                        targetFeatTable = input_targettable)
```

```{r}
annotation_result <- peakPantheR_parallelAnnotation(init_annotation, ncores=0,
                                                    curveModel='skewedGaussian',
                                                    verbose=TRUE)
```

For some reason peakPanther won't annotate these files. The only workaround I have found right now is saving the list of target features as .csv, and loading it back in. See also: https://github.com/phenomecentre/peakPantheR/issues/33

I have honestly no idea how to explain this bug, as both objects "input_targettable" and "inputlist" are identical?


```{r}
write.table(input_targettable, file = "K:/shares/di04_limet_bioinformatics/PhD Pablo/Publicaties/WIP/TARDIS/data/STDtargets.csv")
```

```{r}
inputlist <- read.table("K:/shares/di04_limet_bioinformatics/PhD Pablo/Publicaties/WIP/TARDIS/data/STDtargets.csv")
```

```{r}
init_annotation <- peakPantheRAnnotation(spectraPaths = input_spectraPaths,
                        targetFeatTable = inputlist)
```

```{r}
annotation_result <- peakPantheR_parallelAnnotation(init_annotation, ncores=0,
                                                    curveModel='skewedGaussian',
                                                    verbose=TRUE)
```


```{r}
data_annotation <- annotation_result$annotation
updated_annotation  <- annotationParamsDiagnostic(data_annotation, verbose=TRUE)

```
```{r}
outputAnnotationDiagnostic(updated_annotation, saveFolder="K:/shares/di04_limet_bioinformatics/PhD Pablo/Publicaties/WIP/TARDIS/data/peakpanther_results", 
                            savePlots=TRUE,
                            verbose=TRUE, ncores=2)
```


```{r}
update_csv_path <- 'K:/shares/di04_limet_bioinformatics/PhD Pablo/Publicaties/WIP/TARDIS/data/peakpanther_results/annotationParameters_summary.csv'

# load csv
new_annotation <- peakPantheR_loadAnnotationParamsCSV(update_csv_path)
#> uROIExist set to TRUE
#> New peakPantheRAnnotation object initialised for 2 compounds


new_annotation <- resetFIR(new_annotation)
```

```{r}

new_annotation <- resetAnnotation(new_annotation, spectraPaths= input_spectraPaths,
                                useUROI=TRUE, useFIR=TRUE)
```

```{r}
# annotate files serially
new_annotation_result <- peakPantheR_parallelAnnotation(new_annotation, 
                                                        ncores=1, verbose=FALSE)
```

```{r}
peakTables(new_annotation_result$annotation)
```

```{r}
final_annotation <- new_annotation_result$annotation

outputAnnotationDiagnostic(final_annotation, saveFolder= "K:/shares/di04_limet_bioinformatics/PhD Pablo/Publicaties/WIP/TARDIS/data/peakpanther_results/final_annotation results",
                        savePlots=TRUE,verbose=TRUE)
```

PeakPanther manages to find all 25 peaks in the analytical standard mixtures. However, it's peak filling (fallback integration functionality when no line can be fitted) seems to be a little aggressive, filling in peaks where there is barely any raw data present (more clearly visible on the diagnostic plots). The intensity of those filled peaks is really low though, so it might not be such a big issue if we choose to filter afterwards. Concerning the quality metrics, the curve fitting also gives some weird behavior throughout. 


## Own approach: TARDIS

My own approach is really simple and was created to recreate as much as possible the way targeted peak picking works in Thermo XCalibur.
I've poored the whole thing into one function, sourced from a different script.

```{r}


source(file = "S:/di04_limet_bioinformatics/PhD Pablo/R_code/TARDIS/R/functions/peakdet.R")
source(file = "S:/di04_limet_bioinformatics/PhD Pablo/R_code/TARDIS/R/functions/quality.R")
source(file = "S:/di04_limet_bioinformatics/PhD Pablo/R_code/TARDIS/R/functions/createRanges.R")
source(file = "S:/di04_limet_bioinformatics/PhD Pablo/R_code/TARDIS/R/functions/createTargetList.R")
source(file = "S:/di04_limet_bioinformatics/PhD Pablo/R_code/TARDIS/R/functions/plotQCs.R")
source(file = "S:/di04_limet_bioinformatics/PhD Pablo/R_code/TARDIS/R/functions/plotSamples.R")
source(file = "S:/di04_limet_bioinformatics/PhD Pablo/R_code/TARDIS/R/functions/peaks_with_tardis.R")

start <- Sys.time()

dbData <- createTargetList(input_directory_targets = "D:/Data/Metabease/metabolomics_feces/positive/targets.xlsx",pos_pattern = "+",neg_pattern = "-",polarity = 'positive',ion_column = "ion",columns_of_interest = c("id","name","mz","rt"))

#batch_positions = list(c(1,149),c(150,294))
batch_positions = list(c(1,20),c(21,40))

result <- tardis_peaks(file_path = "D:/Data/Metabease/metabolomics_feces/positive/", dbData = dbData, ppm = 5, rtdev = 18, mode = "metabolomics" , polarity = "positive", output_directory =  "D:/Data/Metabease/metabolomics_feces/positive/", plots_samples = TRUE, plots_QC = TRUE, batch_mode = TRUE, batch_positions = batch_positions, sample_pattern = "", QC_pattern = "QC",int_std_id = c("331","1578","1576","1583","1577"))

end <- Sys.time()

elapsed <- end-start


```


